---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Diagnostic horizons and other differentiae {#sec-diagnostics}

```{r}
library(dplyr)
library(tibble)
library(gt)
library(gtExtras)
```

Diagnostic horizons and other differentiating criteria are defined to facilitate the assignment of soils to classes @tbl-diagsum. The definitions are not intended to represent a comprehensive classification of horizons.

|  |  |
|:-----------------------------------|:-----------------------------------|
| **Horizons, Pans and Layers:** | **Soil materials and Contacts:** |
| [Argillic horizon](#sec-diag-argh) | [Allophanic soil material](#sec-diag-alloph) |
| [Brittle-B horizon](#sec-diag-britb) | [Anthropic Soil Material](#sec-diag-anthro) |
| [Calcareous horizon](#sec-diag-calch) | [Lithic contact](#sec-diag-lithc) |
| [Cutanic horizon](#sec-diag-cuth) | [Organic soil material](#sec-diag-org) |
| [Cutanoxidic horizon](#sec-diag-cutoxh) | [Paralithic contact](#sec-diag-plith) |
| [Densipan](#sec-diag-dens) | [Saline soil material](#sec-diag-salt) |
| [Distinct topsoil](#sec-diag-dts) | [Sulfidic soil material](#sec-diag-sulfd) |
| [Duripan](#sec-diag-dpan) | [Sulfuric soil material](#sec-diag-sulfr) |
| [Eluvial horizon](#sec-diag-eluvh) | [Tephric soil material](#sec-diag-teph) |
| [Fragipan](#sec-diag-fpan) | [Vitric soil material](#sec-diag-vitr) |
| [Humus-pan](#sec-diag-hpan) | **Profile Forms:** |
| [Ironstone-pan](#sec-diag-ipan) | [Gley profile form](#sec-diag-gleypf) |
| [Nodular horizon](#sec-diag-nodh) | [Mottled profile form](#sec-diag-mottpf) |
| [Ortstein-pan](#sec-diag-opan) | **Features:** |
| [Oxidic horizon](#sec-diag-oxh) | [Fluvial features](#sec-diag-fluv) |
| [Peaty topsoil](#sec-diag-pts) | [Perch-gley features](#sec-diag-pgley) |
| [Placic horizon](#sec-diag-plac) | [Sodic features](#sec-diag-sod) |
| [Podzolic-B horizon](#sec-diag-podzb) | **Other Differentiae:** |
| [Redox-mottled horizon](#sec-diag-roxh) | [Childs' Test](#sec-diag-fe2) |
| [Reductimorphic horizon](#sec-diag-redmh) | [Low-chroma colours](#sec-diag-lch) |
| [Slowly permeable layer](#sec-diag-slowp) | [Reactive-aluminium test](#sec-diag-naf) |
| [Weathered-B horizon](#sec-diag-bw) | [Redox segregations](#sec-diag-rsegs) |

: Summary of diagnostic horizons and other differentiae {#tbl-diagsum .hover .sm .borderless}

The following diagnostic horizons and soil characteristics have been modified from Soil Taxonomy @soilsurveystaff1999: [argillic horizon](#sec-diag-argh), [duripan](#sec-diag-dpan), [fragipan](#sec-diag-fpan), [lithic contact](#sec-diag-lithc), [paralithic contact](#sec-diag-plith) and [placic horizon](#sec-diag-plac).

Transitional horizons (e.g. AB or A/B horizons) are not included when considering horizon notations for *New Zealand Soil Classification* criteria unless explicitly stated. Note that:

-   "in some part" means that the requirement must be met in some subhorizon or some subsample that is 10 cm or more thick within the specified thickness.
-   "in the major part" means that the requirement must be met over more than half the specified thickness (e.g. from the base of the A horizon to 60 cm from the mineral soil surface). Analyses or estimates should be used from subhorizons that subdivide the thickness, or if the subhorizons are not recognised, then from subsamples of the relevant horizons.

## Allophanic Soil Material {#sec-diag-alloph .unnumbered}

Allophanic soil material has soil properties dominated by nanocrystalline minerals (structured at the nanometre scale, previously termed short-range order), namely allophane and ferrihydrite. Imogolite, which comprises long, thin hollow nanotubes, and thus has both long- and short-range order, may occur with allophane but usually in small amounts. Other terms such as "amorphous", "poorly crystalline" and "non-crystalline", have been previously used for such minerals but in most cases are no longer valid [@churchman2012]. Their chief characteristics are reactive variable-charge surfaces, and a very high specific surface area (several hundreds of m^2^/g).

Allophanic soil material has BOTH

1.  *Either*
    (a) All of the following (in a fresh sample):
        (i) Sensitive or strongly sensitive sensitivity class (distinctly greasy or smeary feel except in some sandy soils), *and*
        (ii) Very weak or weak unconfined soil strength class (when moist), *and*
        (iii) Non-sticky or slightly sticky stickiness class, *and*
        (iv) Strong or very strong [reactive-aluminium test](#sec-diag-naf); *or*
    (b) P retention of 85% or more; AND
2.  Dry bulk density of the fine-earth fraction (where the volume is determined on a field-moist soil) of less than 0.9 Mg/m^3^.

Layers meeting the requirements of allophanic soil material may also meet the requirements of [vitric soil material](#sec-diag-vitr).

Accessory chemical properties relate to variable-charge characteristics, P retention, and high organic-matter contents. Accessory physical properties include high total available water capacity and readily available water capacity, and low penetration resistance. In addition, allophanic soil material undergoes irreversible changes upon drying, for example, in plastic and liquid limits and in apparent particle-size distribution. It should be noted that minerals other than allophane (e.g. ferrihydrite) can give rise to allophanic soil material.

## Anthropic Soil Material {#sec-diag-anthro .unnumbered}

Anthropic soil material constitutes human-altered or human-transported soil material that shows evidence of the purposeful modification of soil properties or of surface features by human activity. The field evidence of alteration excludes low-impact practices that still allow the soil to meet the requirements of other soil orders (e.g., shallow cultivation, ploughing, the addition of amendments). Artifically drained soils/wetlands are also not included here because it is difficult to unambiguously link artificial drainage to specific, field-identifiable soil indicators. The nature of alteration can comprise profile truncation, mixing, inversion, deposition of transported material, and addition of artefacts. Due to the high variability of the morphological or chemical soil properties that can result from human soil modifications, the criteria of anthropic soil material are mainly linked to their anthropic genesis, and thus deviate from the [principles of the *New Zealand Soil Classification*](Introduction.qmd#sec-intro-principles). Common indicators of anthropic soil material are:

-   excavated or constructed landforms,
-   mechanically detached, redistributed, and re-oriented soil material,
-   abraded and broken rock fragments that do not correspond with the fragments in the underlying soil material (e.g., fractures that cut through rather than between individual minerals),
-   presence of manufactured materials of any kind and age
-   waste products associated with human activities (e.g., food, manufacturing, construction, mining)
-   Cultural/religious activity linked to soil (e.g., burial sites).

The relationship between natural soil horizons or parent materials with anthropic soil material or truncation is commonly in the form of a lithological discontinuity, indicated by abrupt or sharp transitions between horizons. The majority of anthropic soil material (including truncation) in New Zealand is the result of distinct and extensive modification from the mid-19^th^ century coinciding with early European settlement, deforestation, and farming. From the mid-20^th^ century, excavation and construction (e.g., infrastructure, roading, industrial and urban expansion) has increasingly intensified these impacts on our soils. The influence of ‘older’ Māori cultural practice (e.g., soil alteration/modification practices, Māori plaggen soils [@puke2021]) is still recognised in many parts of New Zealand, and can predate European settlement.

## Argillic Horizon {#sec-diag-argh .unnumbered}

An argillic horizon is a clay-enriched horizon. It is indicated by a **Bt** horizon notation as in Btg, BCt, etc. It has ONE of the following:

1.  It is vertically continuous and is 10 cm or more thick. Clay coatings occur that have a waxy lustre when dry and sufficient thickness to envelop fine sand grains. Coatings occur either on peds (10% or more of the ped surfaces), or in pores (in more than one-third of the observed tubular pores) or as bridges between sand grains (in more than half of the horizon); OR
2.  It is composed of clay-enriched lamellae (clay bands), that within 90 cm of the mineral soil surface have a combined thickness of 15 cm or more; OR
3.  It contains sufficiently more clay than the overlying horizon, as detected by hand texturing (5% or more) , excluding any differences which result from a lithological discontinuity, and *either*
    (a) it is overlain by an [eluvial horizon](#sec-diag-eluvh) and the upper boundary of texture contrast is abrupt or sharp, *or*
    (b) clay coatings occur on ped or pore surfaces.

Horizons having coatings which do not meet the requirements of an argillic horizon are likely to meet the requirements of a [cutanic horizon](#sec-diag-cuth).

## Brittle-B Horizon {#sec-diag-britb .unnumbered}

A brittle-B horizon is a B or BC horizon that has ALL of the following:

1.  Brittle failure (the horizon may contain rock fragments, but the fine-earth fraction must be sufficiently coherent to allow brittle failure); AND
2.  It is apedal-massive. Extremely coarse or gross prisms may be present, if the interior of the prisms is apedal-massive; AND
3.  Few or less fine roots occur throughout the horizon.

The brittle-B horizon differs from the [fragipan](#sec-diag-fpan) by having either lower soil strength or lower penetration resistance. Extremely coarse prism faces, if present, are not defined by [low-chroma colours](#sec-diag-lch) as they are in a fragipan. The horizon commonly has some roots throughout, in contrast to the fragipan in which roots are confined to cracks. A brittle-B horizon is usually given the horizon notation suffix **(x)**.

## Calcareous Horizon {#sec-diag-calch .unnumbered}

A calcareous horizon is a horizon in which calcium carbonate occurs in the fine-earth fraction. The concentration may be as low as 1%, but its presence can be detected by any effervescent reaction with 10% HCl on samples from a freshly exposed profile. The calcium carbonate may be inherited from a calcareous parent material ('primary' carbonates), or it may be formed in the soil and occur as coatings on rock fragments, threadlike deposits in pores, or as nodules ('secondary' carbonates). A calcareous horizon is given the notation suffix **k**.

## Childs' test {#sec-diag-fe2 .unnumbered}

This test indicates reduced conditions, manifested by [low-chroma colours](#sec-diag-lch) and/or a mottled profile form because of the presence of exchangeable and water-soluble ferrous iron (Fe^2+^). The reagent 2,2’-bipyridine, also known as ⍺,⍺’-dipyridyl, [@childs1981] was originally applied to the soil as solution, but indicator strips are now also available [@berkowitz2017]. The latter are preferred, but the use of a spray bottle to apply the solution should definitely be avoided due to the toxicity of the reagent. The development of a red or pink colour indicates a positive reaction (i.e., reduced conditions), normally within about 30 minutes [@schoeneberger2012]. Note that in some cases, waterlogged low-chroma-coloured soils may not give positive tests if water-soluble ferrous ions have been leached from the system (i.e., Fe^2+^ is not present in the soil any more).

## Cutanic Horizon {#sec-diag-cuth .unnumbered}

A cutanic horizon is a B or BC horizon containing translocated material forming dark-coloured coatings on ped faces, in pores or on rock fragments. The coatings fail to meet the requirements for coatings of an [argillic horizon](#sec-diag-argh) or a Bh horizon.

It meets BOTH of the following:

1.  The coatings do not have a waxy lustre when dry or are not sufficiently thick to envelop fine sand grains. Silt coatings are excluded. (Silt coatings have similar colour to the matrix, or have higher value and/or lower chroma than the matrix. On drying they may be thick enough to meet [argillic horizon](#sec-diag-argh) requirements, and have flow-like surfaces, but they have a matt rather than a waxy lustre.) AND
2.  The coatings have moist colour values of 4 or less, or value 5 and chroma 3 or less.

Many soils have horizons with coatings on peds or in pores which are very thin, do not have a waxy lustre when dry and have lower colour value than the matrix. It is difficult in the field to decide whether these coatings are inorganic or organic, and whether they are derived by illuviation from overlying horizons, by movement within horizons or by *in-situ* weathering. The [cutanic horizon](#sec-diag-cuth) is designed to recognise such horizons. A cutanic horizon is usually given the horizon notation suffix **(h)**.

## Cutanoxidic Horizon {#sec-diag-cutoxh .unnumbered}

A cutanoxidic horizon [@wilson1987] is a strongly weathered, clayey, low-activity-clay horizon. The dominant clays are kaolin group minerals, and clay coatings occur on less than 10% of ped faces. Exchangeable aluminium, as a percentage of CEC, is usually greater than 25% (and is frequently more than 50% in some part of the horizon). It has ALL the following characteristics:

1.  It meets the requirements of a [cutanic horizon](#sec-diag-cuth), AND
2.  It is a B horizon that is clayey and has polyhedral peds 2-6 mm; AND
3.  Soil materials lack the characteristic friable failure over a wide range in moisture contents that is exhibited by [oxidic horizons](#sec-diag-oxh). The horizon has a failure class of friable only at water contents close to field capacity. Small changes in water content from field capacity result in large changes in soil strength and failure. Semi-deformable failure occurs at water contents wetter than field capacity. Very firm or stronger ped strength with brittle failure occurs at soil water matric potentials drier than about 30 kPa; AND
4.  Soil materials are sticky and very plastic, in comparison to oxidic materials which are only slightly sticky in relation to their clay content; AND
5.  Peds have smooth faces with silt-sized aggregates of iron oxide crystallites which give the faces a dusty appearance when dry. The latter property in particular distinguishes this horizon from horizons developed in well-drained [Brown Soils](#sec-ord-B).

The significance of this horizon lies in the combination of low ECEC, clay illuviation, acidity and physical properties different to that of [oxidic horizons](#sec-diag-oxh).

## Densipan {#sec-diag-dens .unnumbered}

A densipan is a non-cemented [eluvial horizon](#sec-diag-eluvh), denoted with the suffix **d**, of very high soil strength and bulk density. It meets ALL the following requirements:

1.  It is an [eluvial horizon](#sec-diag-eluvh); AND
2.  *Either*
    (a) Unconfined strength is hard or very hard at soil water states from near wet to dry, *or*
    (b) Soil penetration resistance exceeds 4000 kPa at soil water states from near wet to dry; AND
3.  Moist and dry samples slake in water.

Densipans occur in soils with felsic parent materials. The strength is due to a close-fit arrangement of sand and silt-sized quartz particles. It differs from a [duripan](#sec-diag-dpan) by lack of cementation.

## Distinct Topsoil {#sec-diag-dts .unnumbered}

A distinct topsoil [modified from @avery1980] is normally designated an **A** horizon and has BOTH of the following:

1.  Moist colour value and/or chroma is less than that of the horizon below; AND
2.  Thickness is 5 cm or more (including any F, H or O topsoil layer).

The distinct topsoil is used to distinguish minimal soil development in the distinction between [Recent Soils](Key-to-Orders.qmd#sec-ord-R) and [Raw Soils](Key-to-Orders.qmd#sec-ord-W).

## Duripan {#sec-diag-dpan .unnumbered}

A duripan is a subsurface horizon that is cemented by silica or other opaque or uncoloured material. It has ALL of the following requirements, but does not meet the requirements of the [calcareous horizon](#sec-diag-calch):

1.  Dry fragments do not slake in water even during prolonged wetting; AND
2.  It does not react visibly with 10% HCl; AND
3.  The average lateral distance between any fractures is 100 mm or more.

The duripan is recognised in [Pallic Soils](Key-to-Orders.qmd#sec-ord-P) where the cementing materials are apparently related to the presence of tephra in the parent material or high exchangeable sodium in the soil. A duripan is given the horizon notation suffix **q**.

## Eluvial horizon {#sec-diag-eluvh .unnumbered}

An eluvial horizon is a mineral horizon with the horizon notation E that, as a result of downward or lateral movement, contains less organic matter, pedogenic oxides, clay, or a combination of these than the horizon immediately below. The primary notation **E** is used for these horizons. Relative to an overlying **H**, **O** or **A** horizon, it contains less organic matter. The colour of the horizon can be that of uncoated sand and silt grains but eluvial horizons are also recognised where coats of iron oxides or other minerals mask the colour of the primary particles. It has ALL of the following characteristics:

1.  A moist colour value of 4 or more, or a dry colour value of 5 or more, with or without skeletans; AND
2.  *Both*
    (a) A higher colour value or lower chroma combined with less well developed pedality or coarser texture than an underlying B horizon; *and*
    (b) A higher colour value than an overlying **H**, **O** or **A** horizon.

## Fluvial Features {#sec-diag-fluv .unnumbered}

The intention of fluvial features is to recognise soils with parent materials that result from transportation, sorting and deposition by water. Fluvial features are used to differentiate [Recent Soils](Key-to-Orders.qmd#sec-ord-R), [Raw Soils](Key-to-Orders.qmd#sec-ord-W) and parent material classes of soil families or series that occur on landforms formed through fluvial processes.

Relevant landforms include floodplains, estuarine surfaces, lacustrine surfaces, aggregating fan surfaces, levees, backplains, bars, channels, deltas, floodplain benches, outwash plains and swamps (all defined by @obrien2025).

Confirming soil characteristics include:

1.  A buried A horizon or some other field indication of an irregular change in carbon with depth (such as sedimentary plant-leaf material).
2.  Sedimentary bedding in C (or transitional C) horizons, indicating deposition in water (such as scoured surfaces, cross-stratification, sedimentary laminations, current ripples or foreset beds).
3.  An unripened horizon with fluid, or very fluid, fluidity class in some layer with an upper surface within 120 cm of the soil surface.
4.  In [tephric soil materials](#sec-diag-teph)
    a.  disturbance or overthickening of the regional sequence of tephras;
    b.  rounded or subrounded rock fragments;
    c.  presence of non-volcanic rock fragments.

The emphasis here on genetic landform criteria is not consistent with the principle that soils should be classified on the basis of similarity of measurable soil properties rather than presumed genesis (see [Introduction](#sec-intro)). Measurable soil properties that will group together the required soils have not been recognised. The confirmatory soil properties, however, will aid class assignment decisions in many cases.

## Fragipan {#sec-diag-fpan .unnumbered}

A fragipan is an apparently non-cemented subsoil horizon that has high bulk density (usually ≥ 1.5 Mg/m^3^) with high strength when dry. It has ALL of the following:

1.  An air-dried clod must slake when fully immersed in water; AND
2.  Brittle failure when moist (the horizon may be dominated by rock fragments but the fine-earth material must be Sufficiently coherent to allow brittle failure); AND
3.  It has at least slightly firm moist soil strength; AND
4.  *Either*
    (a) Extremely coarse or gross prismatic peds: the prisms have apedal-massive interiors, or break to secondary peds with horizontal dimensions of 100 mm or more, and the prism faces are defined by colours of chroma 3 or less, *or*
    (b) The horizon is apedal-massive throughout, or has extremely coarse or gross prismatic peds, and the moist soil strength is very firm; AND
5.  If roots are present, they are confined predominantly to planar voids between prisms or to worm burrows; AND
6.  Penetration resistance under moist conditions is 3100 kPa or more; AND
7.  It does not occur within an [Eluvial horizon](#sec-diag-eluvh).

A fragipan is given the horizon notation suffix **x**, usually in combination with a B or BC horizon.

## Gley Profile Form {#sec-diag-gleypf .unnumbered}

A gley profile form is defined by the presence of a [reductimorphic horizon](#sec-diag-redmh) with an upper boundary within 30 cm of the mineral soil surface.

Soils with a gley profile form have usually been recognised as poorly or very poorly drained profiles in the soil drainage classification of @obrien2025. Typically, they will give a positive response to [Childs’ test](#sec-diag-fe2). A gley profile form can develop due to impeded vertical drainage of surface waters, lateral subsurface flow or a shallow groundwater table.

## Humus-Pan {#sec-diag-hpan .unnumbered}

A humus-pan is a B horizon that is 10 mm or more thick and is normally given the horizon designation **Bhm**. It has ALL of the following requirements:

1.  It is apedal-massive; AND
2.  It has either firm or stronger moist soil strength, brittle failure when moist, or moist penetration resistance of 3100 kPa or more; AND
3.  It has dominant moist colour value in the matrix of 3 or less, or moist colour value of 4 if the chroma is 2; AND
4.  It contains more than 1.0% organic carbon.

## Ironstone-Pan {#sec-diag-ipan .unnumbered}

An ironstone-pan is an indurated B horizon that is given the notation **Bsm**. It is dominantly composed of iron oxides with or without manganese oxides. It has ALL of the following characteristics:

1.  The upper boundary is distinct, abrupt or sharp; AND
2.  It is weakly or strongly indurated; AND
3.  Fresh fracture surfaces are black and have a metallic lustre; AND
4.  It forms a continuous horizon, or it is fractured into blocks of 100 mm (in horizontal dimension) or more; AND
5.  It is 10 mm or more thick.

Ironstone-pans commonly occur at a textural discontinuity, in the fluctuating zone of a water-table. It is likely that the iron has been precipitated from iron-rich groundwater moving laterally. In Taranaki [@childs1990], the pans are porous and some appear to have formed as iron-oxide rhizomorphs, which have been progressively infilled and welded together by further precipitations of iron oxide. In these pans the mineralogy is dominated by varying proportions of goethite and ferrihydrite.

Ironstone-pans are not usually associated with [eluvial horizons](#sec-diag-eluvh) and do not occur in [Podzols](Key-to-Orders.qmd#sec-ord-Z). They differ from [ortstein-pans](#sec-diag-opan) and [placic horizons](#sec-diag-plac) which are often associated with [eluvial horizons](#sec-diag-eluvh) in [Podzol](Key-to-Orders.qmd#sec-ord-Z) or [Brown Soils](Key-to-Orders.qmd#sec-ord-B), and which have high organic carbon levels. Ironstone pans meet the requirements of the material below a petroferric contact of *Soil Taxonomy*.

The pans are a barrier to plant roots. Heavy machinery is required to break them up for the installation of drains. The permeability of the pans is likely to be slow.

## Lithic Contact {#sec-diag-lithc .unnumbered}

A lithic contact occurs at the contact of soil with underlying rock. The rock is hard or very hard and is impracticable to dig with a spade.

*In situ* rocks in New Zealand are commonly jointed at intervals of less than 100 mm, and consequently the lithic contact definition of *Soil Taxonomy* often fails to apply [@laffan1979]. The lithic contact is defined here as follows:

At a lithic contact, rock fragments accommodate one another with non-random orientation with respect to any geological structure that may be present, and cracks or joints are mostly less than 5 mm wide.

The lithic contact may be subdivided into coherent-lithic or shattered-lithic materials.

*Coherent-lithic* materials are equivalent to materials beneath the lithic contact of *Soil Taxonomy*. Cracks or joints occur at horizontal intervals of more than 100 mm. They often cause drainage impedance.

*Shattered-lithic* materials are similar except that joints or cracks may occur at intervals of less than 100 mm. Shattered-lithic materials differ from fragmental or skeletal materials in which there is no continuity of geological structure between adjacent rock fragments, and rock fragment faces do not accommodate one another. Shattered-lithic materials are more permeable than coherent-lithic materials, and offer a significant rooting volume.

## Low-chroma colours {#sec-diag-lch .unnumbered}

Low-chroma colours are dark, pale, white, or blue-grey ('gley') colours that have a moist chroma of ≤ 2 or a moist chroma of 3 if the value is ≥ 6.

## Mottled Profile Form {#sec-diag-mottpf .unnumbered}

A mottled profile form is defined by EITHER

1.  A [redox-mottled horizon](#sec-diag-roxh) with an upper boundary within 30 cm of the mineral soil surface; OR
2.  A [reductimorphic horizon](#sec-diag-redmh) with an upper boundary between 30 and 60 cm of the mineral soil surface.

Soils with a mottled profile form have usually been recognised as imperfectly drained profiles in the soil drainage classification of @taylor1968. Redox mottles are formed as a result of the reduction and solubilisation of iron and/or manganese, their translocation and concentration, and their oxidation and precipitation in the form of oxides. Mottles that have originated in some other way (e.g. rock colour patterns or skeletans) are excluded. A mottled profile form can develop due to impeded vertical drainage of surface waters, lateral subsurface flow or fluctuations of the groundwater table.

## Nodular Horizon {#sec-diag-nodh .unnumbered}

A nodular horizon has more than 15% (by volume) nodules, as segregations of iron or aluminium oxyhydroxides, with some kaolinite, in a layer more than 10 cm thick. Nodular horizons are given the horizon notation suffix **i**.

The nodules are common features in [Oxidic Soils](Key-to-Orders.qmd#sec-ord-X) and some [Granular Soils](Key-to-Orders.qmd#sec-ord-N) [@wilson1987]. The frequency distribution of nodules is clustered in the \< 2% range and the \> 15% range. Few profiles are known to lie in between.

The nodular horizon limit is intended to exclude thin layers of rewashed nodules on colluvial footslopes, and also infrequent localised concentrations in soils with characteristically few nodules.

## Organic Soil Material {#sec-diag-org .unnumbered}

Organic soil material is soil material dominated by organic matter, excluding fresh litter and living plant material. Organic soil material usually has at least 18% organic carbon (approximately 30% organic matter) but it is defined here using morphology and simple analyses for easier recognition.

Organic soil material has EITHER

1.  *All* of the following:
    (a) Colour value moist of 3 or less (after exposure to air) and colour value dry of 4 or less, *and*
    (b) Deformable failure, *and*
    (c) Weight loss of 65% or more by oven-drying a field-saturated sample; OR
2.  More than 20% (by volume) unrubbed fibre content; OR
3.  More than 35% (by weight) loss on ignition except in materials dominated by [allophanic soil material](#sec-diag-alloph) or by limestone; OR
4.  18% or more total carbon.

Unrubbed fibres are pieces of plant tissue large enough to be retained on a 100-mesh (0.15 mm) sieve, except for wood fragments that cannot be crushed or shredded in the hand and are larger than 2 cm in the smallest dimension. Rubbed fibre is the fibre that remains after rubbing a wet sample 10 times between the thumb and forefinger, or kneading a ball in the palm 10 times using firm pressure.

Organic soil materials that have been accumulated under wet conditions are subdivided into three classes, based on evidence of decomposition [@clayden1989]. These classes are used to distinguish soil groups of [Organic Soils](Key-to-Orders.qmd#sec-ord-O).

*Fibric soil material* (Of horizon) consists mainly of well preserved plant remains that are readily identifiable in terms of botanical origin. The fibre content after rubbing is at least 75% by volume.

*Mesic soil material* (Om horizon) consists mainly of partially decomposed plant remains and does not meet the requirements of either fibric soil material or humified soil material. This material correlates with hemic soil material in other soil classification systems.

*Humified soil material* (Oh horizon) consists of strongly decomposed organic matter with few or no identifiable plant remains other than resistant woody fragments \> 20 mm that cannot be reduced to fibres by crushing and shredding between the fingers. The fibre content is less than 15% after rubbing. This material correlates with sapric soil material defined in other soil classification systems.

## Ortstein-Pan {#sec-diag-opan .unnumbered}

An ortstein-pan is a B horizon that is normally given the horizon notation **Bsm**. It has ALL of the following requirements:

1.  Thickness of more than 10 mm; AND
2.  The upper boundary is sharp or abrupt; AND
3.  It is massive and has either firm or stronger moist soil strength, or has moist penetration resistance of 3100 kPa or more; AND
4.  It does not meet the requirements of a [humus-pan](#sec-diag-hpan), and does not have the metallic lustre of fresh fracture surfaces of an [ironstone-pan](#sec-diag-ipan).

Ortstein-pans develop from precursor [podzolic-B horizons](#sec-diag-podzb) into cemented horizons. They are normally associated with an overlying [eluvial horizon](#sec-diag-eluv) and underlying podzolic-B or [weathered-B](#sec-diag-bw) horizons. They typically occur in [Podzol Soils](#sec-ord-Z). They are thicker than [placic horizons](#sec-diag-plac) and constitute a root barrier if continuous.

## Oxidic Horizon {#sec-diag-oxh .unnumbered}

The oxidic horizon is a strongly weathered B horizon consisting of mixed crystalline iron and aluminium oxides and kaolin group minerals, with low activity clay properties. An oxidic horizon is given the horizon notation **Bo**. It has ALL of the following requirements:

1.  Weak or very weak primary ped strength and soil strength as determined by the unconfined resistance-to-crushing test at moist to dry soil water states; AND
2.  Unconfined failure is friable or very friable over very moist to dry soil water states. Materials fail to predominantly 3 mm or smaller peds comprising silt- and sand-sized polyhedra and spheroids; AND
3.  Primary peds slake rapidly in water to stable microaggregates (3 mm or smaller) which show no dispersion or slight dispersion; AND
4.  Non-reactive or very weakly reactive to the [reactive-aluminium test](#sec-diag-naf), AND
5.  Materials are at most slightly sticky and plastic.

Oxidic horizons are clayey, with measured clay contents commonly exceeding 60%. The measured clay percentage is usually larger than in overlying A horizons, but clay increase is not a defining criterion because of the problem of quantifying clay contents in materials that are frequently difficult to disperse.

lay coatings are either visually absent or only present at frequencies of about 1%. The oxidic horizon has low activity clay accessory properties with ECEC and CEC less than 12 and 16 cmol~c~/kg clay respectively.

## Paralithic Contact {#sec-diag-plith .unnumbered}

A paralithic contact is the upper surface of rock or regolith material that has ALL of the following requirements:

1.  It can be cut with difficulty with a spade; AND
2.  Wet penetration resistance exceeds 2600 kPa; AND
3.  Roots if present are few and confined to cracks; AND
4.  If the overlying horizon is a [reductimorphic](#sec-diag-redmh) or [redox-mottled](#sec-diag-roxh) horizon, low chroma or high chroma mottles are less common below the contact.

The paralithic contact meets the definition of a paralithic contact of *Soil Taxonomy*, but without the restrictive requirement for spacing of cracks. The horizon beneath the contact is given the horizon designation CR. Paralithic contacts may occur either on weakly weathered or unweathered rocks which are not strongly lithified, or on saprolites which have become soft by strong weathering.

## Peaty Topsoil {#sec-diag-pts .unnumbered}

A peaty topsoil is 10 cm or more thick and is saturated for 30 or more consecutive days in most years (unless it is artificially drained), and has EITHER

1.  More than 30% organic matter (i.e., peat or peaty textures), OR
2.  Textures with peat (17–30% organic matter) if the clay content is less than 18%.

In some subgroups a peaty topsoil may be buried by a surface layer of new material of up to 60 cm in thickness.

## Perch-Gley Features {#sec-diag-pgley .unnumbered}

Perch-gley features are the morphologic indicators of saturation and reducing conditions caused by a water-table perched on a [slowly permeable layer](#sec-diag-slowp) within the soil profile.

A horizon with perch-gley features EITHER

1.  Has [redox-segregations](#sec-diag-rsegs) that occur mainly within peds, or in the case of an apedal soil, mainly within the soil matrix. Macro-void surfaces, either partings or pores, are dominated by [low-chroma colours](#sec-diag-lch). Iron and manganese precipitates occur either adjacent to the low-chroma void surfaces as a coating or as discrete mottles within the soil mass; OR
2.  Overlies a horizon that has fewer [redox-segregations](#sec-diag-rsegs) and is not a [reductimorphic horizon](#sec-diag-redmh).

## Placic Horizon {#sec-diag-plac .unnumbered}

A placic horizon is a thin iron pan that is normally designated **Bfm**. It has ALL of the following:

1.  It is 10 mm or less thick; AND
2.  It is at least weakly indurated, and is black to reddish brown or dark red in colour. A black upper part can often be distinguished from a reddish brown lower part; AND
3.  The upper and lower boundaries are sharp, and may be smooth, wavy or convolute in shape.

The placic horizon usually occurs as a single pan but in places can be bifurcated. It is equivalent to the placic horizon of *Soil Taxonomy* except that New Zealand iron pans are enriched in iron and organic matter without significant accumulations of manganese [@clayden1990].

## Podzolic-B Horizon {#sec-diag-podzb .unnumbered}

A podzolic-B horizon is a B horizon that meets ONE of the following:

1.  It meets the requirement of a Bh horizon (because it has colour value and chroma of 3 or less, or value 4 and chroma 2, dominant in the matrix, and contains more than 1% organic carbon). The fabric has sand- or silt-size pellet-like aggregates, coats on mineral grains, or both; OR
2.  *Both*
    (a) It is associated with an overlying (but not necessarily immediately overlying) [eluvial horizon](#sec-diag-eluvh) in which weathered films on sand and silt particles are either absent, very thin or discontinuous, so that the colour of the horizon is mainly determined by the colours of the uncoated grains. The moist colour value of the [eluvial horizon](#sec-diag-eluvh) is 4 or more (or a dry colour value is 5 or more). It has higher colour value or lower chroma and less well developed pedality than an underlying B horizon; *and*
    (b) The B horizon (or some part of it) is 5 cm or more thick and meets the requirements of a Bs (or Bs(g) or Bs(f)) horizon because it has a strong [reactive-aluminium test](#sec-diag-naf), and at least *one* of the following:
        (i) Reddish hue and highest chroma at the top of the horizon; *or*
        (ii) Earthy structure, or weakly developed blocks or polyhedra; *or*
        (iii) Very weak or weak soil strength when moist or dry; *or*
        (iv) Sand- or silt-sized pellet-like aggregates; OR
3.  It meets the definition of a Bs horizon (part 2(b) above) and has in addition, coatings of value 4 or less *either*
    (a) On 50% or more ped faces, *or*
    (b) As patches covering 20% or more of cut faces.

## Reactive-Aluminium Test {#sec-diag-naf .unnumbered}

This test indicates the presence of reactive hydroxy-aluminium groups, as occur for example in allophane and aluminium-humus complexes [@obrien2025].

Using the procedure of @fieldes1966, 1 drop of saturated sodium fluoride (NaF) solution is placed on a small test sample of soil, which has been smeared on to a filter paper treated with phenolphthalein indicator. The soil sample must be moist. For classification, the reactivity of the soil sample is placed into one of classes in @tbl-ft-naf.

::: {.content-visible unless-format="epub"}
```{r}
#| label: tbl-ft-naf
#| tbl-cap: "Sodium fluoride reaction ratings"

dat_ft_naf <-
  tribble(~Code, ~Name, ~Description, ~Colour, 
          'N', 'No reaction',  'No colour change', '#FFFFFF',
          'L', 'Slight reaction', 'Paper turns faint pink within two minutes',  '#FFBADE',
          'M', 'Moderate reaction', 'Paper turns moderate pink, usually within 30 seconds', '#CF3A69',
          'H', 'Strong reaction', 'Paper turns dark pink, usually instantly', '#A11844'
  ) |>
  dplyr::mutate(fa_icon = 'fas fa-circle')

tbl_ft_naf <- gt(dat_ft_naf) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  fmt_icon(columns    = fa_icon, 
           fill_color = from_column("Colour"), 
           height     = '1.5em',
           fill_alpha = 0.9) |>
  cols_hide(Colour) |>
  cols_label(fa_icon = 'Colour') |>
  cols_width(Code ~pct(10), Name ~ pct(20), fa_icon ~ pct(10))
  
tbl_ft_naf
```
:::

::: {.content-visible when-format="epub"}
```{r}
#| label: tbl-ft-naf2
#| tbl-cap: "Sodium fluoride reaction ratings"

dat_ft_naf <-
  tribble(~Code, ~Name, ~Description, ~Colour, 
          'N', 'No reaction',  'No colour change', '',
          'L', 'Slight reaction', 'Paper turns faint pink within two minutes',  '',
          'M', 'Moderate reaction', 'Paper turns moderate pink, usually within 30 seconds', '',
          'H', 'Strong reaction', 'Paper turns dark pink, usually instantly', ''
  )

tbl_ft_naf <- gt(dat_ft_naf) |>
  tab_options(
    column_labels.font.weight = 'bold', 
    heading.title.font.weight = 'bold',
    table.align = 'center',
    table.width = '85%'
  ) |>
  tab_style(style = list(cell_text(weight = 'bold',
                                   align  = 'center')),
            locations = cells_body(columns = 1)) |>
  ## colour in Col 4
  tab_style(style = list(cell_fill(color = '#FFFFFF')),
                         locations = cells_body(columns = 4, rows = 1)) |>
  tab_style(style = list(cell_fill(color = '#FFBADE')),
                         locations = cells_body(columns = 4, rows = 2)) |>
  tab_style(style = list(cell_fill(color = '#CF3A69')),
                         locations = cells_body(columns = 4, rows = 3)) |>
  tab_style(style = list(cell_fill(color = '#A11844')),
                         locations = cells_body(columns = 4, rows = 4)) |>
  cols_width(Code ~pct(10), Name ~ pct(20), Colour ~ pct(10))
  
tbl_ft_naf
```
:::

## Redox-Mottled Horizon {#sec-diag-roxh .unnumbered}

A redox-mottled horizon is a horizon affected in parts by reducing conditions as indicated by the presence of [redox-segregations](#sec-diag-rsegs). These usually indicate intermittent saturation of the soil by water. A redox-mottled horizon is identified by the suffix **(f)** or **(g)**.

A redox-mottled horizon has 2% or more redox segregations. If [low chroma colours](#sec-diag-lch) occur, they must occupy less than 50% of the matrix exposed in a cut face of the horizon and must not be dominant on ped faces.

The intermittent wetness may be caused by intermittent perched water, or by the fluctuating upper limits of deeper more prolonged groundwater. A redox-mottled horizon may represent more prolonged saturation and reduction in parent materials that are predominantly andesitic or basaltic compared with other parent materials.

## Redox Segregations {#sec-diag-rsegs .unnumbered}

Redox segregations are mottles or concretions formed as a result of the reduction and solubilisation of iron and/or manganese, their translocation, concentration, and their oxidation and precipitation in the form of oxides @clayden1989. They may have low or high chroma colours, or both.

The nature of the water table is indicated by the association of low and high chroma colours. If subject to reduction by perched water, the [low chroma colours](#sec-diag-lch) are likely to be at ped or pore surfaces and the high chroma colours are likely to be within the soil matrix. If the soil is subject to reduction by groundwater, the low chroma colours are likely to be within the soil matrix and the high chroma colours are likely to be at ped or pore surfaces. Reducing conditions may also be indicated by the presence of sufficient ferrous iron to give a positive reaction to [Childs' test](#sec-diag-fe2).

## Reductimorphic Horizon {#sec-diag-redmh .unnumbered}

A reductimorphic horizon has a texture class with peat, or [low chroma colours](#sec-diag-lch) that occupy 50% or more of the matrix exposed in a cut face of the horizon or are dominant on ped faces. A reductimorphic horizon includes any subjacent horizons with peaty texture class.

A reductimorphic horizon is a horizon strongly affected by reducing conditions as indicated by [low-chroma colours](#sec-diag-lch) consistent with long saturation by water, and will give a postive response to [Childs' test](#sec-diag-fe2). The prolonged wetness may be caused by a water-table perched on a [slowly permeable layer](#sec-diag-slowp) within the soil profile or by a groundwater-table. A reductimorphic horizon is identified by the horizon suffix **g** or **r**.

## Saline soil material {#sec-diag-salt .unnumbered}

Soil salinity refers to the accumulation of water-soluble salts, mainly of sodium, calcium and magnesium. These can severely affect plant growth and land use, and increase soil erosion [@craw2024].

Most soils generate some salt as a product of chemical weathering, with composition and concentration determined by the parent material lithology. Soils may also receive additional salt at their surface, transported by wind and/or rain. These salts accumulate as crystals in the soil profile or on the soil surface when rainfall is too low to dissolve and remove them. Some soils are affected by saline groundwater from marine or hydrothermal sources.

Soil salinity is somewhat difficult to measure in the field in ways that can be strongly related to plant growth [@derm2011]. As a compromise, the electrical conductivity (EC) of a 1:5 solution of soil to deionised water is measured. Results exceeding 0.8 mS/cm within in the top 60 cm of the soil profile are considered significant enough to be considered in the classification of [Gley](#sec-ord-G), [Recent](#sec-ord-R), [Raw](#sec-ord-W) and [Semi-Arid](#sed-ord-S) soil orders.

Note that salinity at any given location (including within the profile) is expected to be highly variable across seasons and will generally be at its maximum towards the end of a dry summer.

## Slowly Permeable Layer {#sec-diag-slowp .unnumbered}

A soil horizon that functions as a slowly permeable layer is one in which the vertical saturated hydraulic conductivity is less than 4 mm/h (1.0 × 10^-6^ m/s) as measured by a standard method. If no measurement is available then the horizon can be identified by the following morphological characteristics [@griffiths1991]:

A slowly permeable layer meets ONE of the following requirements:

1.  *Both*
    (a) The soil material is pedal; *and*
    (b) More than half of the peds are coarser than 10 mm (mean of the x and y axes in a horizontal plane) and meet *one* of the following:
        (i) Peds 20 to 50 mm, with degree of packing at least extremely high; *or*
        (ii) Peds 50 to 100 mm with degree of packing at least very high; *or*
        (iii) Peds 100 mm or larger with degree of packing at least high; OR
2.  *Either*
    (a) The soil texture class is sand or loamy sand and has an extremely high degree of packing; *or*
    (b) The soil material has a texture group other than sandy and has at least a high degree of packing.

A slowly permeable layer is significant for the land use, genetic and hydrological understanding of soils. Other diagnostic horizons that frequently also function as slowly permeable layers are [fragipans](#sec-diag-fpan), [argillic horizons](#sec-diag-argh), [densipans](#sec-diag-dens), [humus-pans](#sec-diag-hpan) or [ortstein-pans](#sec-diag-opan) and [lithic](#sec-diag-lithc) or [paralithic contacts](#sec-diag-plith).

## Sodic Features {#sec-diag-sod .unnumbered}

A horizon with sodic features, identified by the horizon notation suffix **n**, has significant exchangeable sodium and is not necessarily characterised by a high soluble salt content. It has BOTH

1.  *Either*
    (a) Exchangeable sodium percentage of 6% or more (or exchangeable sodium is 0.7 cmol~c~/kg or more); *or*
    (b) When a 10 mm diameter sample which has been air-dried is placed in distilled water or salt free water a cloud of dispersed clay will form within 10 minutes around the sample. This test will not apply if the soil is cemented; AND
2.  *Either*
    (a) Clay or clay/organic coatings have colour value 4 or less; *or*
    (b) Prismatic or blocky peds; *or*
    (c) It may be overlain by an Ew, Ew(g) or Ew(f) horizon, or have skeletans (visible on dry ped faces) near the top of the horizon.

## Sulfidic soil material {#sec-diag-sulfd .unnumbered}

Sulfidic soil materials contain more than 0.01% (dry mass) inorganic sulfide sulfur, mainly in the form of pyrite (FeS~2~) as very fine crystals. These sulfides are stable under anaerobic conditions but will otherwise rapidly oxidise, releasing sulfuric acid. Their field pH in water will usually be neutral to mildly alkaline, but pH in hydrogen peroxide may drop as low as 2 after a vigorous reaction [@obrien2025]. These paired field pH tests may be used as an initial pre-screening test for inorganic sulfides, but are not definitive on their own. In the field, sulfidic soil material will be waterlogged and will meet the criterium of a [reductimorphic horizon](#sec-diag-redmh) with a hue of 5Y or less red, or black if organic material or metasulfides are dominant.

Sulfidic soil materials can be split into hyper- and hypo- sulfidic subclasses. Methods for determining these subclasses may be found in @sullivan2018.

*Hypersulfidic soil material* has a high acidification risk. These materials become [sulfuric soil material](#sec-diag-sulfr) when disturbed.

*Hyposulfidic soil material* has a lower acidification risk due to the presence of substances that buffer the acidity released upon oxidation of sulfides (e.g., carbonates). These materials will not become [sulfuric soil material](#sec-diag-sulfr) when disturbed, but may affect water quality by mobilisation of metals.

Sulfidic soil materials are most commonly silt- or clay-textured, but can still be sandy or even gravelly under specific depositional conditions. They sometimes contain elevated organic matter and may even comprise marine-influenced peats. Sulfidic soil materials are predominantly found near the land surface in low-energy coastal environments and buried at depth under floodplains. They require anaerobic conditions, active sulfate-reducing bacteria, a supply of iron and sulfate, and enough groundwater movement to remove reaction products and sustain the activity of the iron sulfide accumulation process [@rijkse1978].

Note that there is potential for sulfidic soil material to underly soils of soil orders that do not currently contain sulfidic subgroups. This can apply to any soil in a near-coastal depositional landscape at low elevation (\<20 m as guidance), provided the required conditions are present. This material will most often be buried deeper than 1 m. These deeply buried materials remain out of scope for the classification.

## Sulfuric soil material {#sec-diag-sulfr .unnumbered}

Sulfuric soil materials are extremely acidic as a result of inorganic sulfide oxidation releasing large amounts of sulfuric acid. Sulfuric soil materials have ALL of the following properties:

1.  Under field conditions have *either*
    (a) a pH in water of \<4.0 (if mineral soil material), *or*
    (b) a pH in water of \<3.0 (if [organic soil material](#sec-diag-orgm)), AND
2.  At least *one* of the following,
    (a) concentrations of sulfate minerals, notably the mineral jarosite with a distinctive butter-yellow colour; *or*
    (b) directly overlying [sulfidic soil material](#sec-diag-sulfd), *or*
    (c) 0.05% or more sulfate sulfur (by dry mass).

Sulfuric soil material will initially have a similar appearance to the [sulfidic soil material](#sec-diag-sulfd) from which they develop, but will generally show redox segregations. As oxidation continues, the matrix colour progresses from [low-chroma colours](#sec-diag-lch) to yellow-brown and red hues. Physical shrinking and irreversible cracking may also occur in material with high clay content.

## Tephric Soil Material {#sec-diag-teph .unnumbered}

Tephric soil material occurs in or below the soil solum. It includes:

*Tephra:* The unconsolidated, primary pyroclastic (fragmental) products of explosive volcanic eruptions encompassing all grain sizes (i.e. ash: grains \<2 mm in diameter; lapilli: 2–64 mm; blocks/bombs: \>64 mm), and all compositions (rhyolitic, andesitic, or basaltic), irrespective of emplacement mechanism (Lowe 2011). Tephra thus includes unconsolidated deposits derived from non-welded (i.e., not lithified) pyroclastic flows (density currents) or surges as well as widespread fall deposits. Tephras comprise three main components: volcanic glass, including glass shards, pumice (highly vesicular glass), and scoriae; crystals (mineral grains); and lithics (rock fragments) [@lowe2017].

*Tephra deposits:* The material derived at least partly from tephra that has been reworked and mixed with material from other sources. They include tephric loess, tephric blown sand and volcanogenic alluvium. As a general guide, tephric deposits from andesitic sources have more than 10% volcanic glass in the sand fraction and those from rhyolitic sources have more than 40% volcanic glass in the sand fraction.

Tephric soil material may include soil materials that meet the requirements of [allophanic soil material](#sec-diag-alloph) or [vitric soil material](#sec-diag-vitr). It is used to distinguish soil groups of the [Raw Soils](#sec-ord-W) and [Recent Soils](#sec-ord-R), and parent material classes at soil family or series level.

## Vitric Soil Material {#sec-diag-vitr .unnumbered}

Vitric soil material [@parfitt1984] has more than 35% rock fragments (2 mm or greater, by volume) of which at least 60% is rhyolitic pumice, or there is more than 40% sand of which more than 30% is volcanic glass or crystals coated with glass [@eden1992].

## Weathered-B Horizon {#sec-diag-bw .unnumbered}

A weathered-B horizon shows evidence of chemical alteration of the original parent material it is derived from and is indicated by a **Bw** horizon notation as in Bw, Bw(g), Bw(f), etc. It has at least ONE of the following:

1.  Redder hue or higher chroma than an underlying horizon in similar materials; OR
2.  Have spheroidal, blocky, polyhedral, tabular, prismatic, columnar or platy pedality which distinguish the horizon from a BC or C horizon below; OR
3.  Evidence of either partial or complete decalcification, i.e. less CaCO~3~ than the underlying horizon which may contain redeposited carbonates.

A weathered-B horizon may also meet the requirements of a [redox-mottled horizon](#sec-diag-roxh), [argillic horizon](#sec-diag-argh), [cutanic horizon](#sec-diag-cuth), or [brittle-B horizon](#sec-diag-britb).

{{< pagebreak >}}
